---
title: "Norwegian CDI"
author: "Colin Ayres"
output: html_document
date: "December 10, 2024"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We read data from "norwegian_cdi.csv" file.
We only keep the "WG" subset (there are two age versions,
"**W**ords & **G**estures" for younger and "**W**ords & **S**entences" for older children).
We remove columns we don't need, keeping only
"child_id", "age", "comprehension", "production", "sex", and "caregiver_education".

```{r data_import}
read.csv2("norwegian_cdi.csv") -> cdi
cdi[cdi$form == "WG",] -> cdi  
subset(cdi, select = c(child_id, age, comprehension, production, sex, caregiver_education)) -> cdi
```

There are `r nrow(cdi)` responses in total,
with age ranging from `r min(cdi$age)` to `r max(cdi$age)` months.

Below is a scatterplot of *production* scores by age.
Blue dots are female scores and pink dots are male scores.

```{r plot}
# "col" argument (below) can take a single value, e.g., "red",
# or a whole vector of values, e.g., specifing colour for each dot.
plot(cdi$age, cdi$production,
     col = ifelse(cdi$sex == "Male", "pink", "blue"),
     xlab = "Months", ylab = "Production score")
```

We focus now on the oldest group. There are `r nrow(cdi[cdi$age==20,])` children in this group.
The bounds of the 95% CI for the true (population) mean of *comprehension* scores for them is printed
below:

```{r oldest-CI95-comprehension}
mean(cdi[cdi$age==20,]$comprehension, na.rm = TRUE) -> xbar
sd(cdi[cdi$age==20,]$comprehension, na.rm = TRUE) -> sd
nrow(cdi[cdi$age==20,]) -> n
xbar + qt(c(.025, 0.975), n-1) * sd / sqrt(n) -> cdi20_comprehension
cdi20_comprehension
```

The bounds of the 95% CI for the true (population) mean of *production* scores for them is printed
below:

```{r oldest-CI95-production}
mean(cdi[cdi$age==20,]$production, na.rm = TRUE) -> xbar
sd(cdi[cdi$age==20,]$production, na.rm = TRUE) -> sd
nrow(cdi[cdi$age==20,]) -> n
xbar + qt(c(.025, 0.975), n-1) * sd / sqrt(n) -> cdi20_production
cdi20_production
```


## Bonus

We also calculate the *comprehension* CIs for girls and boys (in the oldest group).

> It's not obligatory! Don't worry, if you can't work out how to use `tapply()` below,
> but please remember to set `eval` to `FALSE` for the chunk in that case,
> so that the document knits even without working code below.

```{r oldest-CI95-comprehension-sex, eval=TRUE}
# tapply() splits a vector given as the first argument,
# using vector(s) given as the second argument,
# and to each part of the original vector applies a function
# given as the third argument.
tapply(cdi[cdi$age==20,]$comprehension, cdi[cdi$age==20,]$sex, function(x) {
    mean(x) + qt(c(0.025, 0.975), length(x) - 1) * sd(x) / sqrt(length(x))
})
```

