---
title: "Norwegian CDI"
author: ""
output: html_document
date: ""
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

> Fill in the `author` and `date` fields in the heading and delete this line.

We read data from "norwegian_cdi.csv" file.
We only keep the "WG" subset (there are two age versions,
"**W**ords & **G**estures" for younger and "**W**ords & **S**entences" for older children).
We remove columns we don't need, keeping only
"child_id", "age", "comprehension", "production", "sex", and "caregiver_education".

```{r data_import}
read.csv2("norwegian_cdi.csv") -> cdi
cdi[cdi$form == "WG", ] -> cdi
cdi[, c("child_id", "age", "comprehension", "production", "sex", "caregiver_education")] -> cdi
```

There are `r nrow(cdi)` responses in total,
with age ranging from `r min(cdi$age)` to `r max(cdi$age)` months.

Below is a scatterplot of *production* scores by age.
Blue dots are female scores and pink dots are male scores.

```{r plot}
# "col" argument (below) can take a single value, e.g., "red",
# or a whole vector of values, e.g., specifing colour for each dot.
plot(cdi$age, cdi$production,
     col = ifelse(cdi$sex == "Male", "pink", "blue"),
     xlab = "Months", ylab = "Production score")
```

We focus now on the oldest group. There are `r sum(cdi$age == 20)` children in this group.
The bounds of the 95% CI for the true (population) mean of *comprehension* scores for them is printed
below:

```{r oldest-CI95-comprehension}
cdi[cdi$age == 20,] -> cdi20
mean(cdi20$comprehension) +
     qt(c(.025, .975), nrow(cdi20) - 1) * sd(cdi20$comprehension) / sqrt(nrow(cdi20))
```

The bounds of the 95% CI for the true (population) mean of *production* scores for them is printed
below:

```{r oldest-CI95-production}
mean(cdi20$production) +
     qt(c(.025, .975), nrow(cdi20) - 1) * sd(cdi20$production) / sqrt(nrow(cdi20))
```


## Bonus

We also calculate the *comprehension* CIs for girls and boys (in the oldest group).

> It's not obligatory! Don't worry, if you can't work out how to use `tapply()` below,
> but please remember to set `eval` to `FALSE` for the chunk in that case,
> so that the document knits even without working code below.

```{r oldest-CI95-comprehension-sex, eval=TRUE}
# tapply() splits a vector given as the first argument,
# using vector(s) given as the second argument,
# and to each part of the original vector applies a function
# given as the third argument.

tapply(cdi20$comprehension, cdi20$sex, function(x) {
    mean(x) + qt(c(.025, .975), length(x) - 1) * sd(x) / sqrt(length(x))
})
```


## Merging

```{r item-data-import, eval = FALSE}
read.csv("norwegian_cdi_items.csv") -> items
```


```{r merging, eval = FALSE}
merge(cdi20, items, all.y = FALSE) -> items_extra

# table(items_extra$child_id)
# mean(table(items_extra$child_id))
# length(table(items_extra$child_id))
# mean(items_extra$age)
```

