---
title: "Norwegian CDI"
author: "Colin Ayres"
output: html_document
date: "December 10, 2024"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We read data from "norwegian_cdi.csv" file.
We only keep the "WG" subset (there are two age versions,
"**W**ords & **G**estures" for younger and "**W**ords & **S**entences" for older children).
We remove columns we don't need, keeping only
"child_id", "age", "comprehension", "production", "sex", and "caregiver_education".

```{r data_import}
# Lines below contain various fragments of required commands
read.csv("norwegian_cdi.csv") -> cdi
"WG"  -> cdi
"child_id", "age", "comprehension", "production", "sex", "caregiver_education" -> cdi
```

There are `r ` responses in total,
with age ranging from `r ` to `r ` months.

Below is a scatterplot of *production* scores by age.
Blue dots are female scores and pink dots are male scores.

```{r plot}
# "col" argument (below) can take a single value, e.g., "red",
# or a whole vector of values, e.g., specifing colour for each dot.
?(, ,
     col = ifelse(),
     xlab = "Months", ylab = "Production score")
```

We focus now on the oldest group. There are `r sum()` children in this group.
The bounds of the 95% CI for the true (population) mean of *comprehension* scores for them is printed
below:

```{r oldest-CI95-comprehension}
 -> cdi20

```

The bounds of the 95% CI for the true (population) mean of *production* scores for them is printed
below:

```{r oldest-CI95-production}

```


## Bonus

We also calculate the *comprehension* CIs for girls and boys (in the oldest group).

> It's not obligatory! Don't worry, if you can't work out how to use `tapply()` below,
> but please remember to set `eval` to `FALSE` for the chunk in that case,
> so that the document knits even without working code below.

```{r oldest-CI95-comprehension-sex, eval=TRUE}
# tapply() splits a vector given as the first argument,
# using vector(s) given as the second argument,
# and to each part of the original vector applies a function
# given as the third argument.

tapply(, , function(x) {
    mean(x) + qt(c( , ), ) *
})
```

